Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-13 write-ups/PLDI/paper.tex

\begin{lstlisting}
(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * pools) {
(*@\label{line:wksteal:getgroupid}@*)  int pool_id = get_group_id();
  while (more_work(pools)) {
(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(pools, pool_id);
    if (t) {
      // task processing may create more work
(*@\label{line:wksteal:processtask}@*)      process_task(t, pools, pool_id);
    }
  }
}
\end{lstlisting}

@@

\paragraph{Work stealing example}
%
The work stealing paradigm is useful when a computation can be split
into multiple tasks that can be processed independently, but where the
number of tasks is not known ahead of time because processing a given
task may create an arbitrary number of new tasks. Work stealing enables
dynamic balancing of tasks across several processing units, OpenCL
workgroups in our case. Each workgroup has an associated task pool from
which it obtains tasks to process, and to which it stores newly created
tasks. When a workgroup's pool becomes empty, a workgroup---rather than
terminating---tries to \emph{steal} a task from an other
pool. Workgroups keep on processing tasks until all the work is
done. Previous work addresses the feasibility and efficiency of work
stealing on GPUs~ref\cite{dlb-web,TPO10}.

\myfiglong\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue. All threads start their
execution at the function marked with the $\mathsf{kernel}$ keyword
(line~\ref{line:wksteal:kernelfunc}). They receive a pointer to the task
pools, which reside in global memory, and which were initialized by the
host program beforehand to contain the initial tasks. Each thread can
identify the workgroup it belongs to via the function $\getgroupid()$
(line~\ref{line:wksteal:getgroupid}). Here the group id is stored into
the variable $\mathsf{pool\_id}$ (which is, by default, in private
memory) and used to access the relevant task pool. The
$\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
is called by all threads of a workgroup.  Inside
$\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
also called the \emph{master thread} of the workgroup, will try to pop a
task from the workgroup's pool, or steal a task from a different
pool. The master thread uses local memory inside
$\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
all threads of its workgroup. If a task was obtained, then all threads
of the workgroup participate in the processing of the task via the
$\mathsf{process\_task}$ function (line~\ref{line:wksteal:processtask}).
Processing this task may lead to the creation of further tasks, which
the master thread will push to the workgroup's pool.

Although not depicted here, concurrent accesses to each pool inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex, implemented using atomic compare and swap operations on
global memory.  As a result, the kernel presents two opportunities for
spin-waiting: spinning to obtain a mutex, and spinning in the main
kernel loop to obtain a task to process.  The algorithm thus depends
on a fair scheduler, which, as discussed in \mysec\ref{sec:intro},
is not guaranteed by current GPU programming models.

@@

For the work stealing example, there is no state to transmit since a
computation is entirely parameterised by a task, which is retrieved from
a task pool located in global memory. In the main loop, before trying to
obtain a task, we make the workgroup offer itself to be forked or killed
(lines~\ref{line:cwksteal:offerfork},\ref{line:cwksteal:offerkill}). Note
that a workgroup may be killed even if its associated task pool is not
empty, since remaining tasks could eventually be stolen by other
workgroups.  Care must be taken on computing pool identifiers: the
original version in \myfig\ref{fig:workstealing} can rely on the fact
that there are as many pools as workgroups, but here the number of
workgroups may change during execution. Therefore, the number of pools
must now be explicitely given as an argument
(line~\ref{line:cwksteal:argnumpool}) and used as an upper bound of pool
indexes when scanning pools in $\mathsf{more\_work}$
(line~\ref{line:cwksteal:morework}) and $\mathsf{pop\_or\_steal}$
(line~\ref{line:cwksteal:poporsteal}). Besides, computation of the
workgroup's pool index (line~\ref{line:cwksteal:getgroupid}) must not
only be bounded by the number of pools but also be moved inside the main
loop since a forked workgroup that starts to execute after $\offerfork$
must re-compute its pool index.

@@

\begin{lstlisting}
kernel work_stealing(global Task * pools,
(*@\label{line:cwksteal:argnumpool}@*)                     const int num_pools) {
(*@\label{line:cwksteal:morework}@*)  while (more_work(pools, num_pools)) {
(*@\label{line:cwksteal:offerfork}@*)    offer_fork();
(*@\label{line:cwksteal:offerkill}@*)    offer_kill();
(*@\label{line:cwksteal:getgroupid}@*)    int pool_id = get_group_id() % num_pools;
(*@\label{line:cwksteal:poporsteal}@*)    Task * t = pop_or_steal(pools, pool_id, num_pools);
    if (t) {
      // task processing may create more work
(*@\label{line:cwksteal:processtask}@*)      process_task(t, pools, pool_id);
    }
  }
}
\end{lstlisting}

\caption{Cooperative version of the work stealing example from \myfig\ref{fig:workstealing}; the number of workpools
must be explicitly passed as arguments since the number of workgroups
is not a constant anymore}\label{fig:workstealing-cooperative}
\end{figure}

>>>>

\begin{lstlisting}
(*@\label{line:wksteal:kernelfunc}@*)kernel work_stealing(global Task * queues) {
(*@\label{line:wksteal:getgroupid}@*)  int queue_id = get_group_id();
  while (more_work(queues)) {
(*@\label{line:wksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id);
    if (t) {
      // task processing may create more work
(*@\label{line:wksteal:processtask}@*)      process_task(t, queues, queue_id);
    }
  }
}
\end{lstlisting}

@@

\paragraph{Work stealing example}
%
The work stealing paradigm is useful when a computation can be split
into multiple tasks that can be processed independently, but where the
number of tasks is not known ahead of time because processing a given
task may create an arbitrary number of new tasks. Work stealing
enables dynamic balancing of tasks across several processing units,
OpenCL workgroups in our case. Each workgroup has an associated task
queue from which it obtains tasks to process, and to which it stores
newly created tasks. When a workgroup's queue becomes empty, the
workgroup---rather than terminating---tries to \emph{steal} a task
from an other queue. Workgroups keep on processing tasks until all the
work is done. Previous work addresses the feasibility and efficiency
of work stealing on GPUs~ref\cite{dlb-web,TPO10}.

\myfiglong\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue. All threads start their
execution at the function marked with the $\mathsf{kernel}$ keyword
(line~\ref{line:wksteal:kernelfunc}). They receive a pointer to the task
queues, which reside in global memory, and which were initialized by the
host program beforehand to contain the initial tasks. Each thread can
identify the workgroup it belongs to via the function $\getgroupid()$
(line~\ref{line:wksteal:getgroupid}). Here the group id is stored into
the variable $\mathsf{queue\_id}$ (which is, by default, in private
memory) and used to access the relevant task queue. The
$\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
is called by all threads of a workgroup.  Inside
$\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
also called the \emph{master thread} of the workgroup, will try to pop a
task from the workgroup's queue, or steal a task from a different
queue. The master thread uses local memory inside
$\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
all threads of its workgroup. If a task was obtained, then all threads
of the workgroup participate in the processing of the task via the
$\mathsf{process\_task}$ function (line~\ref{line:wksteal:processtask}).
Processing this task may lead to the creation of further tasks, which
the master thread will push to the workgroup's queue.

Although not depicted here, concurrent accesses to queues inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex per queue, implemented using atomic compare and swap operations
on global memory.  As a result, the kernel presents two opportunities
for spin-waiting: spinning to obtain a mutex, and spinning in the main
kernel loop to obtain a task to process.  The algorithm thus depends
on a fair scheduler, which, as discussed in \mysec\ref{sec:intro}, is
not guaranteed by current GPU programming models.

@@

In this example there is no state to transmit since a computation is
entirely parameterised by a task, which is retrieved from a queue
located in global memory. In the main loop, before trying to obtain a
task, we make the workgroup offer itself to be forked or killed
(lines~\ref{line:cwksteal:offerfork},\ref{line:cwksteal:offerkill}). Note
that a workgroup may be killed even if its associated task queue is
not empty, since remaining tasks could eventually be stolen by other
workgroups.  Care must be taken on computing queue identifiers: the
original version in \myfig\ref{fig:workstealing} can rely on the fact
that there are as many queues as workgroups, but here the number of
workgroups may change during execution. Therefore, the number of
queues must now be explicitely given as an argument
(line~\ref{line:cwksteal:argnbqueue}) and used as an upper bound of
queue indexes when scanning queues in $\mathsf{more\_work}$
(line~\ref{line:cwksteal:morework}) and $\mathsf{pop\_or\_steal}$
(line~\ref{line:cwksteal:poporsteal}). Besides, computation of the
workgroup's queue index (line~\ref{line:cwksteal:getgroupid}) must now
be moved inside the main loop since a forked workgroup that starts to
execute after $\offerfork$ must re-compute its group id.

@@

\begin{lstlisting}
kernel work_stealing(global Task * queues,
(*@\label{line:cwksteal:argnbqueue}@*)                     const int nb_queues) {
(*@\label{line:cwksteal:morework}@*)  while (more_work(queues, nb_queues)) {
(*@\label{line:cwksteal:offerfork}@*)    offer_fork();
(*@\label{line:cwksteal:offerkill}@*)    offer_kill();
(*@\label{line:cwksteal:getgroupid}@*)    int queue_id = get_group_id();
(*@\label{line:cwksteal:poporsteal}@*)    Task * t = pop_or_steal(queues, queue_id, nb_queues);
    if (t) {
      // task processing may create more work
(*@\label{line:cwksteal:processtask}@*)      process_task(t, queues, queue_id);
    }
  }
}
\end{lstlisting}

\caption{Cooperative version of the work stealing kernel of
  \myfig~\ref{fig:workstealing}, using $\offerfork$ and $\offerkill$,
  and passing the total number of queues
  explicitely}\label{fig:workstealing-cooperative}
\end{figure}
